---
title: "Comparison of Hierarchical Parent Proportion Tests"
author: "Adam Chan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=9)
knitr::opts_knit$set(root.dir = "C:\\Users\\acha5354\\Desktop\\New Folder\\{SW}\\PhD\\Projects\\200518 - Mass Cytometry Analyses")
```

# Data Loading
The following datasets were used in this analysis:

CyTOF:

- Age Chronic Inflammation, Shen-Orr 2016 <br>
1036209 cells from 29 samples
Predicting young/old from 

- Anti-CTLA-4 and Anti-PD-1, Subrahmanyam 2018 <br>
7264780 cells from 24 samples
Predicting response/nonresponse to treatment

- Anti-PD-1, Kreig 2018 <br>
85715 cells from 20 samples
Predicting response/nonresponse to treatment

- Breast Cancer, Wagner 2019 <br>
855914 cells from 194 samples
Predicting tumor/non-tumor breast cancer samples

- CMV, Davis 2019 <br>
18153877 cells from 69 samples
Predicting positive/negative result from CMV titer in flu vaccine patients

- COVID-19 Whole blood CyTOF, Geanon 2021
4747543 cells from 21 hospitalized covid-19 patients and healthy controls 

Flow Cytometry:

- AML, FlowCAP-II 2013 <br>
55711535 cells from 359 samples
Predicting aml/normal patients

- COVID-19 PBMC, Humblet-Baron 2020 <br>
6510011 cells from 49 patients (6 health, 23 mild-to-moderate, 20 severe) samples
Predicting mild vs severe COVID-19 through PBMC immune profile

- COVID-19 CD4 & CD8 T Cells, De Biasi 2020 <br>
Subset of 5000 cells from 12 Healthy controls and 19 covid patients

- COVID-19 CD8 Non Naive, Mathew 2020 <br>
54 healthy donors and 114 covid patients. Analysed 60% of the original
cells which amounted to 11,591,741. Predicting between recovered donors
and healthy donors

- MS HSCT, Ma 2020 <br>
Predicting between relapse and non-relapse patients recovering from
stem cell transplant

scRNA-seq:
- Melanoma, Sade-Feldman 2019

```{r}
#' Parent proportion test scripts
#' This script is to summarise results from the treeParentProportionTest.R Scripts

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(gridExtra)
  library(latex2exp)
})

treeSigFiles <- list.files("data/benchmarking_data/significance",
                                     full.names = TRUE)

load(file = treeSigFiles[1])

tree_sig_df_combined <- tree_sig_df 

for (i in 2:length(treeSigFiles)){
  load(file = treeSigFiles[i])
  
  tree_sig_df_combined <- tree_sig_df_combined %>%
    bind_rows(tree_sig_df)
}

#' ======================================================================
#' Comparison Visualisations
#' ----------------------------------------------------------------------

gg_theme <- theme_bw() +
  theme(panel.border = element_blank(),
        axis.line = element_line(color = 'black'))

cbp1 <- c("#e69f00", "#56b4e9", "#009e73", "#f0e442",
          "#cc79a7", "#0072b2", "#d55e00", "#999999",
          "#f27c7c", "#279a98", "#9a3f44", "#a53093",
          "#4c5e66")

```

# Visualisations

## Comparing proportion relative to parent vs relative to all
```{r}
#' Q-Q plots by ordering parent vs all
qq_df <- tree_sig_df_combined %>%
  dplyr::filter(statAll != 0) %>%
  mutate(neglogp_all = -log(pvalAll, base=10)) %>%
  arrange(dataset, hierarchy, neglogp_all) %>%
  group_by(dataset, hierarchy) %>%
  mutate(row_id = 1:n()) %>%
  select(dataset,hierarchy, row_id, neglogp_all) %>%
  left_join(
    (tree_sig_df_combined %>%
      dplyr::filter(statAll != 0) %>%
      mutate(neglogp_parent = -log(pvalParent, base=10)) %>%
      arrange(dataset, hierarchy, neglogp_parent) %>%
      group_by(dataset, hierarchy) %>%
      mutate(row_id = 1:n()) %>%
      select(dataset,hierarchy, row_id, neglogp_parent)),
    by=c("dataset"="dataset", "hierarchy"="hierarchy", "row_id"="row_id")
    )

ggplot(qq_df %>%
         dplyr::filter(hierarchy %in% c("average", "hopach") &
                  !(dataset %in% c("COVID-19 PBMC, Mathew 2020 - CD8nonNaive", 
                                   "COVID-19 PBMC, Humblet-Baron 2020",
                                   "COVID-19 T cells, De Biasi 2020 - CD8 - home_hosp",
                                   "Anti-CTLA-4 and Anti-PD-1, Subrahmanyam 2018 - PD1 - home_hosp"))) %>%
         mutate(hierarchy = case_when(
           hierarchy == "average" ~ "Average-linkage Hierarchical Clustering",
           hierarchy == "hopach" ~ "Hopach",
           TRUE ~ hierarchy
         ))) +
  geom_point(aes(x=neglogp_all, y=neglogp_parent, color=hierarchy),
             size=1.5) +
  # geom_line(aes(x=neglogp_all, y=neglogp_parent, color=hierarchy))+
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  facet_wrap(~dataset, scales="free", ncol=4) +
  theme_bw() +
  theme(strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(x=TeX("-log_{10}(p-value) Proportion relative to All"),
       y=TeX("-log_{10}(p-value) Proportion relative to Parent"),
       col="Hierarchy Type") # Exported as 8 x 13 for paper

```

