---
title: "Visualisation"
author: "Adam Chan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    number_sections: yes
    self_contained: yes
    theme: lumen
    toc: yes
    toc_depth: 3
    # toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width=10, fig.height=9, message=FALSE)
```

# Data Preprocessing

Original data and .fcs files can be found at: https://www.immport.org/shared/study/SDY478

```{r eval=FALSE}
library(CATALYST)

library(flowCore)
library(flowViz)
library(ggplot2)
library(SingleCellExperiment)

library(dplyr)
library(tidyr)
library(janitor)
library(readxl)

### Get sample information ---------------------------------
samp_dat <- read.table("CMV, Davis 2019\\data\\SDY478-DR34_Subject_2_CyTOF_result.txt", sep="\t") %>%
  row_to_names(row_number = 1) %>%
  clean_names() %>%
  drop_na()

biosample <- read.table("CMV, Davis 2019\\data\\biosample.txt", sep="\t") %>%
  row_to_names(row_number = 1)

lab_test <- read.table("CMV, Davis 2019\\data\\lab_test.txt", sep="\t") %>%
  row_to_names(row_number = 1)

fcs_md <- samp_dat %>%
  left_join(
    lab_test %>% 
      dplyr::filter(NAME_REPORTED == "CMV Ab") %>%
      mutate(CMV_STATUS = ifelse(RESULT_VALUE_REPORTED > 1, "positive", "negative")) %>%
      left_join(biosample %>%
                  select(BIOSAMPLE_ACCESSION, SUBJECT_ACCESSION),
                by="BIOSAMPLE_ACCESSION"),
    by=c("subject_accession" = "SUBJECT_ACCESSION")
    ) %>%
  dplyr::filter(!is.na(CMV_STATUS)) %>%
  clean_names() %>%
  dplyr::rename(condition = cmv_status)

## Read in data
# Only read in files for samples with associated CMV_STATUS
fcsFiles <- list.files('CMV, Davis 2019\\data\\CyTOF_result',
                       pattern=".fcs",
                       recursive = TRUE,
                       full.names = TRUE)[list.files('CMV, Davis 2019\\data\\CyTOF_result',
                                                     pattern=".fcs",
                                                     recursive = TRUE) %in% fcs_md[["file_name"]]]

fcs_raw <- read.flowSet(fcsFiles, 
                        transformation = FALSE, 
                        truncate_max_range = FALSE)

## Subsample cells
fcs_raw2 <- fsApply(fcs_raw, function(x){
  set.seed(12)
  samp_ids <- sample.int(nrow(x), nrow(x)*0.75)
  x[samp_ids,]
})

### Manually gate for singlets ------------------------------
# Attempting to follow https://static-content.springer.com/esm/art%3A10.1038%2Fs41591-019-0381-y/MediaObjects/41591_2019_381_MOESM1_ESM.pdf
biexpTrans <- biexponentialTransform(a=50, b=1.45, c=50, d=1, f=0, w=0)

## Gate cells
autoplot(transform(fcs_raw2[[1]],
                   `(Ce140)Dd`=biexpTrans(`(Ce140)Dd`),
                   `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`)), 
                   "(Ce140)Dd", 
                   "(Ir191)Dd",
         bins = 200)

recGate1<-rectangleGate("(Ce140)Dd"=c(-0.5,1.5),"(Ir191)Dd"=c(2,3.45))
filters1<-filters(list(recGate1))
xyplot(`(Ir191)Dd` ~ `(Ce140)Dd`
       ,data=transform(fcs_raw2[[1]], 
                       `(Ce140)Dd`=biexpTrans(`(Ce140)Dd`),
                       `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`))
       ,filter=filters1
       ,stat=TRUE
       ,margin=FALSE
       ,checkName=FALSE
)    

fcs_raw3_filt <- flowCore::filter(transform(fcs_raw2,
                                            `(Ce140)Dd`=biexpTrans(`(Ce140)Dd`),
                                            `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`)),
                                  recGate1)

fcs_raw3 <- Subset(fcs_raw2, fcs_raw3_filt)

## Gate intact cells
autoplot(transform(fcs_raw3[[1]], 
                   `(Ir193)Dd`=biexpTrans(`(Ir193)Dd`),
                   `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`)), 
         "(Ir193)Dd", 
         "(Ir191)Dd",
         bins = 200)

recGate1<-rectangleGate("(Ir193)Dd"=c(2.75,3.25),"(Ir191)Dd"=c(2.4,3.))
filters1<-filters(list(recGate1))
xyplot(`(Ir191)Dd` ~ `(Ir193)Dd`
       ,data=transform(fcs_raw3[[1]], 
                       `(Ir193)Dd`=biexpTrans(`(Ir193)Dd`),
                       `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`))
       ,filter=filters1
       ,stat=TRUE
       ,margin=FALSE
       ,checkName=FALSE
)    

fcs_raw3_filt <- flowCore::filter(transform(fcs_raw3,
                                            `(Ir193)Dd`=biexpTrans(`(Ir193)Dd`),
                                            `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`)),
                                  recGate1)

fcs_raw3 <- Subset(fcs_raw3, fcs_raw3_filt)

## Gate intact singlets
autoplot(transform(fcs_raw3[[1]], 
                   `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`)), 
         "Cell_length", 
         "(Ir191)Dd",
         bins = 200)

recGate1<-rectangleGate("Cell_length"=c(0,52),"(Ir191)Dd"=c(2.4,3.))
filters1<-filters(list(recGate1))
xyplot(`(Ir191)Dd` ~ `Cell_length`
       ,data=transform(fcs_raw3[[1]], 
                       `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`))
       ,filter=filters1
       ,stat=TRUE
       ,margin=FALSE
       ,checkName=FALSE
)    

fcs_raw3_filt <- flowCore::filter(transform(fcs_raw3,
                                            `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`)),
                                  recGate1)

fcs_raw3 <- Subset(fcs_raw3, fcs_raw3_filt)

## Gate live intact cells
biexpTrans2 <- biexponentialTransform(a=20, b=1, c=1, d=1, f=0, w=0)
autoplot(transform(fcs_raw3[[1]], 
                   `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`),
                   `(In115)Dd`=biexpTrans2(`(In115)Dd`)), 
         "(In115)Dd", 
         "(Ir191)Dd",
         bins = 200)

recGate1<-rectangleGate("(In115)Dd"=c(-2,-1),"(Ir191)Dd"=c(2.4,3.))
filters1<-filters(list(recGate1))
xyplot(`(Ir191)Dd` ~ `(In115)Dd`
       ,data=transform(fcs_raw3[[1]], 
                       `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`),
                       `(In115)Dd`=biexpTrans2(`(In115)Dd`))
       ,filter=filters1
       ,stat=TRUE
       ,margin=FALSE
       ,checkName=FALSE
)    

fcs_raw3_filt <- flowCore::filter(transform(fcs_raw3, 
                                            `(Ir191)Dd`=biexpTrans(`(Ir191)Dd`),
                                            `(In115)Dd`=biexpTrans2(`(In115)Dd`)),
                                  recGate1)

fcs_raw3 <- Subset(fcs_raw3, fcs_raw3_filt)
# sum(fsApply(fcs_raw3, nrow))
# sum(fsApply(fcs_raw2, nrow))

### Get marker information ----------------------------------
pregating_channels <- c("Bead", "DNA1", "DNA2", "Dead", "Cell_length")
lineage_channels <- c("CD57", "CD19", "CD4", "CD8", "IgD", "CD11c", "CD16", "CD3", 
                      "CD38", "CD27", "CD14", "CXCR5", "CCR7", "CD45RA", "CD20", 
                      "CD127", "CD33", "CD28", "CD161", "TCRgd", "CD123", "CD56", 
                      "HLADR", "CD25")
instrument_channels <- c("Time", "Cell_length")

fcs_panel <- pData(parameters(fcs_raw[[1]])) %>%
  select(name, desc) %>%
  mutate(marker_class = ifelse(desc %in% lineage_channels, 
                               "type", 
                               ifelse(desc %in% pregating_channels | is.na(desc), 
                                      "none",
                                      "state")))

### Compile SingleCellExperiment ---------------------------
#' Applied arcsinh transform with cofactor 5
sce <- prepData(x=fcs_raw3, panel=fcs_panel, md=fcs_md,
                cofactor = 5,
                panel_cols = list(channel="name",
                                  antigen="desc"),
                md_cols = list(file="file_name",
                               id="subject_accession",
                               factors=c("condition", "race", "gender", "subject_age", 
                                         "result_value_reported"))
)

assay(sce, "counts") <- NULL # To save some space

### Apply FlowSOM clustering ------------------------------
sce <- cluster(sce, features = "type",
               xdim = 10, ydim = 10, maxK = 20,
               verbose = FALSE, seed = 1)

save(sce, file="data/sce_tomic_cmv_gated.RData")
```



```{r}
suppressPackageStartupMessages({
  library(treekoR)
  library(SingleCellExperiment)
  library(ggtree)
  library(dplyr)
  
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(ggsignif)
})

data_folder <- "CMV, Davis 2019"
data_panel <- ""

pos_class_name <- "positive"
neg_class_name <- "negative"

# Plot marker densities
gg_theme <- theme_bw() +
  theme(panel.border = element_blank(),
        axis.line = element_line(color = 'black'))

cbp1 <- c("#e69f00", "#56b4e9","#cc79a7", "#009e73", "#f0e442",
           "#0072b2", "#d55e00", "#999999",
          "#f27c7c", "#279a98", "#9a3f44", "#a53093",
          "#4c5e66")
```

# Run treekoR

```{r}
#' Load Data
#' Load SCE with FlowSOM Clustering 
#' -------------------------------------------------------------------
load(file="data/sce_tomic_cmv_gated.RData")

exprs <- t(assay(sce, "exprs"))
clusters <- colData(sce)$cluster_id
classes <- colData(sce)$condition
samples <- as.character(colData(sce)$sample_id)
```


```{r}
hierarchy_meths <- c("hopach")

print("Generating hierarchical trees and testing...")

tree_res_list <- lapply(hierarchy_meths,
       function(x) {
           clust_tree <- getClusterTree(exprs,
                             clusters,
                             hierarchy_method=x,
                             scale_exprs=TRUE)
            
            tested_tree <- treekoR::testTree(phylo=clust_tree$clust_tree,
                      clusters=clusters,
                      samples=samples,
                      classes=classes,
                      pos_class_name=NULL)
            
            return(list(clust_tree=clust_tree, tested_tree=tested_tree))
       })

tree_sig_df <- lapply(tree_res_list, 
       function(x) {
           x$tested_tree$data
       }) %>%
    bind_rows() %>%
    mutate(hierarchy = rep(hierarchy_meths, times=unlist(lapply(tree_res_list, 
                                                                    function(x) {nrow(x$tested_tree$data)})))) %>%
    dplyr::select(node, label, isTip, stat_total, stat_parent, pval_total, pval_parent,
                    parent, clusters, hierarchy) %>%
    dplyr::rename(c("parent_node"="parent", "child_leaf_nodes"="clusters")) %>%
    mutate(parent_leaf_nodes = .[match(.[,"parent_node"] %>% pull, 
                            .[,"node"] %>% pull), 
                            "child_leaf_nodes"] %>% pull) %>% 
    mutate(label = ifelse(is.na(label),
                            node,
                            label)) %>%
    mutate(dataset = paste0(data_folder, " - ", data_panel))

print("Calculating proportions to parent and absolute proportions...")

prop_df_list <- lapply(c("hopach", "average"),
       function(x) {
           clust_tree <- getClusterTree(exprs,
                             clusters,
                             hierarchy_method=x,
                             scale_exprs=TRUE)
            
            prop_df <- getCellProp(phylo=clust_tree$clust_tree,
                       clusters=clusters,
                       samples=samples,
                       classes=classes)
            
            return(list(prop_df=prop_df))
       })
```

```{r eval=F}
plotInteractiveHeatmap(tree_res_list[[1]]$tested_tree,
                       clust_med_df = tree_res_list[[1]]$clust_tree$median_freq,
                       clusters=clusters)
```

## Figure 2a

```{r}
# Plot scatterplot with parent vs. all proportions
max_val <- max(abs(c(tree_res_list[[1]]$tested_tree$data$stat_total, tree_res_list[[1]]$tested_tree$data$stat_parent)),
               na.rm=T)

g1 <- ggplot(tree_res_list[[1]]$tested_tree$data %>% # Use node name where label is null
               mutate(label = ifelse(is.na(label),node,label)),
             aes(x = stat_total, y = stat_parent, shape=isTip, col=isTip,
                 data_id=label, tooltip=label))+
  geom_point() + 
  # geom_point(size=1.5)+
  geom_hline(yintercept = 0, linetype="dashed")+
  geom_vline(xintercept = 0, linetype="dashed")+
  coord_equal(xlim=c(-max_val,max_val),ylim=c(-max_val,max_val))+
  labs(x = "Statistic: %total",y = "Statistic: %parent",
       title = "Significance between patient condition using %parent vs %total",
       subtitle = paste0("Dataset - ", 
                         data_folder,
                         ifelse(exists("data_panel"),
                                paste0(" - ", data_panel),
                                ""))) +
  theme_bw() +
  theme(panel.border = element_blank(),
        axis.line = element_line(color = 'black'),
        legend.position = "top") +
  scale_color_manual(values=c("grey10", "grey50"))

# Export as 6x8
g1
```


## Figure 2b

The two populations of interest are cluster 42 (sibling: cluster 54) and cluster 122 (sibling: cluster 64). Their relation to CMV titre result is picture below

```{r}
prop_df_list[[1]]$prop_df %>%
  select(contains(c("_42", "_122")), class, sample_id) %>%
  gather(variable, value, -c(class, sample_id)) %>%
  filter(variable %in% c("perc_total_42", "perc_total_122",
                         "perc_parent_1_42", "perc_parent_1_122")) %>%
  # mutate(variable = factor(variable, levels = apply(expand.grid(c("prop_all", "prop_parent"), c("_19","_60","_126")), 1, paste0, collapse=""))) %>%
  mutate(variable = factor(variable, 
                           levels = sapply(c("perc_total", "perc_parent_1"), 
                                           function(x) paste0(x, c("_42", "_122"))) %>% 
                             as.vector)) %>%
  ggplot(aes(x=class, y=value, col=class, fill=class)) +
  geom_boxplot(outlier.size=-1, width=0.3) +
  geom_jitter(alpha=0.75, width=0.2) +
  facet_wrap(~variable, scales="free", nrow=2) +
  gg_theme +
  geom_signif(
    color="grey40",
    comparisons = list(c("positive", "negative")),
    margin_top=0.15,
    map_signif_level = function(p) sprintf("p = %.2g", p),
    test = "t.test"
  ) +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.background =element_blank(),
        axis.line = element_line(size = 0.35),
        legend.position = "bottom") +
  scale_color_manual(values = cbp1) +
  scale_fill_manual(values = alpha(cbp1,0.2)) +
  labs(title="Comparison between proportion to all and proportion to parent",
       subtitle="Proportions measured relative to parent cells per sample",
       fill="Patient Outcome", 
       col = "Patient Outcome",
       y="Proportion")
```


For each of the panels, the densities of the tranformed (arcsinh) marker expression for each cluster of interest have been plotted. The density marked "whole" represents the marker expression of all the cells combined, and the other densities represent the marker expression for all the cells in the corresponding cluster.

# Cluster Marker Expression Density

## Cluster 42
```{r}
set.seed(12)
idx <- sample.int(ncol(sce), 1e5)

clust_vals <- c("42", "54")

cells_data <- exprs %>%
  as.data.frame() %>%
  mutate(cluster = clusters) %>%
  slice(idx)
```


```{r fig.width=10, fig.height=10}
# clust_1, clust_2 are a single or vector value of strings containing the clusters of interest
# clust_2 is the sibling of clust_1
plotFacettedRidgelinePlot <- function(clust_1, clust_2) {
  ### Density Ridgeline Plot
  cells_data %>%
    mutate(population = "whole") %>%
    bind_rows(cells_data %>%
                dplyr::filter(cluster %in% clust_1) %>%
                mutate(population = stringr::str_interp("cluster_${clust_1}")),
              cells_data %>%
                dplyr::filter(cluster %in% c(clust_2)) %>%
                mutate(population = stringr::str_interp("cluster_${clust_2} (${clust_1} sibling)"))) %>%
    select(-cluster) %>%
    pivot_longer(!population, names_to="variable", values_to="value") %>% 
    ggplot() + 
    ggridges::geom_density_ridges(aes(x=as.numeric(value), col=population, fill=population, y=population)) + 
    facet_wrap(~variable, scales="free") +
    # facet_wrap(~variable) +
    labs(title=stringr::str_interp("Density of Cluster ${clust_1} vs. the rest")) +
    scale_color_manual(values = cbp1) +
    scale_fill_manual(values = alpha(cbp1,0.2)) +
    gg_theme +
    theme(strip.background = element_blank(),
          axis.line.y = element_blank(),
          axis.text.y = element_blank())
}

plotFacettedRidgelinePlot("42", "54")
```


## Cluster 122 (72 & 73)

Note: The sibling cluster of interest is cluster 122 which is 72 & 73 combined) and the sibling is just clusters 64


```{r fig.width=10, fig.height=10}
plotFacettedRidgelinePlot(c("72","73"), "64")
```

# Session Info
```{r}
sessionInfo()
```